Scaling rates of true polar wander in convecting planets and moons

### 2017-05-01 tpw_rate.tex
Turn around a couple of sentences.

\begin{equation}
\frac{d \mathbf{H}}{dt} + \mitbf{\Omega} \times \mathbf{H} = 0
\label{eq:euler}
\end{equation}
where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
In this case, Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
For a slowly convecting fluid, such as a planetary mantle, the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
\begin{equation}
\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0.
\label{eq:quasistatic_liouville}
\end{equation}
Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for the principal values of $\mathbf{I}(t)$;  the eigenvectors define the principal axes and the eigenvalues correspond to the principal moments of inertia. The rotation axis is expected to align with the axis for the largest moment.  
In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and the associated surface deflections.
Here we are interested in those contributions that arise from mantle convection.

For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{sabadini1981pleistocene, spada1992excitation}:
\begin{equation}
I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + E_{ij}(t)
\label{eq:separation}
\end{equation}
where $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
If we plug this decomposition into Equation~\eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
\begin{equation}
\mathbf{\Omega} \times (\mathbf{J} \cdot \mitbf{\Omega}) = -\mitbf{\Omega} \times ( \mathbf{E} \cdot \mitbf{\Omega} ).
\label{eq:disequilibrium}
\end{equation}
This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches
of the convective part of the moment of inertia ($\mathbf{E}$) and the rotational deformation part of the moment of inertia ($\mathbf{J}$).
Our goal is to characterize this balance from a perspective of scaling and fluid dynamics.

>>>>

\begin{equation}
\frac{d \mathbf{H}}{dt} + \mitbf{\Omega} \times \mathbf{H} = 0
\label{eq:euler}
\end{equation}
where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
On a dynamic planet $\mathbf{I}$ may be a function of time, so $\mitbf{\Omega}$ must be also vary with time to conserve angular momentum.
In this case, Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
For a slowly convecting fluid, such as a planetary mantle, the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
\begin{equation}
\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0.
\label{eq:quasistatic_liouville}
\end{equation}
Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for the principal values of $\mathbf{I}(t)$;  the eigenvectors define the principal axes and the eigenvalues correspond to the principal moments of inertia. The rotation axis is expected to align with the axis for the largest moment.  
In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and the associated surface deflections.
Here we are interested in those contributions that arise from mantle convection.

The moment of inertia tensor is commonly separated into three parts for mantle convection problems \citep{sabadini1981pleistocene, spada1992excitation}:
\begin{equation}
I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + E_{ij}(t)
\label{eq:separation}
\end{equation}
where $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
If we plug this decomposition into Equation~\eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
\begin{equation}
\mathbf{\Omega} \times (\mathbf{J} \cdot \mitbf{\Omega}) = -\mitbf{\Omega} \times ( \mathbf{E} \cdot \mitbf{\Omega} ).
\label{eq:disequilibrium}
\end{equation}
This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches
of the convective part of the moment of inertia ($\mathbf{E}$) and the rotational deformation part of the moment of inertia ($\mathbf{J}$).
Our goal is to characterize this balance from a perspective of scaling and fluid dynamics.
